---
title: "Introduction"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    lib_dir: libs
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"


---

class: title-slide

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
]

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(gridExtra)
options(scipen = 999)
```

---

class: middle

## Goals


- Linear Regression with Numeric Predictors
- Conditions for Linear Regression

---

class: middle

```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(openintro)
library(broom)
theme_set(theme_gray(base_size = 18))
```

#### Data `babies` in `openintro` package

```{r echo = FALSE}


glimpse(babies)

```

---

##  Baby Weights

.pull-left[
```{r eval = FALSE, fig.align='center', message = FALSE, warning = FALSE}


ggplot(babies, 
       aes(x = gestation, y = bwt)) +
  geom_point()

```

]

.pull-right[

```{r echo = FALSE, fig.align='center', message = FALSE, fig.height= 6,warning = FALSE}


ggplot(babies, 
       aes(x = gestation, y = bwt)) +
  geom_point()

```

]
---

##  Baby Weights

.pull-left[
```{r eval = FALSE, fig.align='center', message = FALSE, warning = FALSE}
ggplot(babies,
         aes(x = gestation, y = bwt)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) 

```

`lm` stands for linear model  
`se` stands for standard error
]

.pull-right[

```{r echo = FALSE, fig.align='center', message = FALSE, fig.height= 6,warning = FALSE}


ggplot(babies,
         aes(x = gestation, y = bwt)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) 

```

]



---

class: middle

<div align = "center">

| y | Response    | Birth weight | Numeric |
|---|-------------|-----------------|---------|
| x | Explanatory | Gestation           | Numeric |


---

## Linear Equations Review


.pull-left[

Recall from your previous math classes

$y = mx + b$

where $m$ is the slope and $b$ is the y-intercept

e.g. $y = 2x -1$
]

--



.pull-right[
```{r echo = FALSE, fig.height = 5, message = FALSE}
x <- c(0, 1, 2, 3, 4, 5)
y <- c(-1, 1, 3, 5, 7, 9)

as.data.frame(x = x, y = y) %>% 
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  scale_y_continuous(breaks = c(-1, 1, 3, 5, 7, 9)) +
  scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5)) +
  geom_smooth(method = "lm", se = FALSE)

```

Notice anything different between baby weights plot and this one?
]

---

class: middle 

.pull-left[

**Math** class

$y = b + mx$

$b$ is y-intercept  
$m$ is slope  
]


.pull-left[

**Stats** class

$y_i = \beta_0 +\beta_1x_i + \epsilon_i$

$\beta_0$ is y-intercept  
$\beta_1$ is slope  
$\epsilon_i$ is error/residual  
$i = 1, 2, ...n$ identifier for each point
]

---

class: middle


```{r}
model_g <- lm(bwt ~ gestation, data = babies)

```

`lm` stands for linear model. We are fitting a linear regression model. Note that the variables are entered in y ~ x order.

---

class: middle

```{r}
broom::tidy(model_g)
```


--

$\hat {y}_i = b_0 + b_1 x_i$

$\hat {\text{bwt}_i} = b_0 + b_1 \text{ gestation}_i$

$\hat {\text{bwt}_i} = -10.1 + 0.464\text{ gestation}_i$

---

class: middle

## Expected bwt for a baby with 300 days of gestation

$\hat {\text{bwt}_i} = -10.1 + 0.464\text{ gestation}_i$

$\hat {\text{bwt}} = -10.1 + 0.464 \times 300$

$\hat {\text{bwt}} =$ `r -10.1 + 0.464*300`


For a baby with 300 days of gestation the expected birth weight is `r -10.1 + 0.464*300` ounces.

---

## Interpretation of estimates

.pull-left[
```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies %>% 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) 
    
```

$b_1 = 0.464$ which means for one unit(day) increase in gestation period the expected increase in birth weight is 0.464 ounces.

]

--

.pull-right[
```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies %>% 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlim(0, 360) +
  ylim(-10, 180) +
  geom_abline(slope = 0.459, intercept = -8.76)
  
  
```

$b_0 = -10.1$ which means for gestation period of 0 days the expected birth weight is -10.1 ounces!!!!!!!!
(does NOT make sense)
]

---

class: middle

## Extrapolation

- There is no such thing as 0 days of gestation.

--

- Birth weight cannot possibly be -10.1 ounces.

--

- Extrapolation happens when we use a model outside the range of the x-values that are observed. After all, we cannot really know how the model behaves (e.g. may be non-linear) outside of the scope of what we have observed. 

---

## Baby number 148

.pull-left[

```{r}
babies %>% 
  filter(case == 148) %>% 
  select(bwt, gestation)

```

]

.pull-right[

```{r echo = FALSE, message = FALSE, fig.height=5, warning = FALSE}

baby_148 <- subset(babies, case == 148)

babies %>% 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = baby_148, color = "red")
```


]

---

class: middle



## Baby #148

.pull-left[

**Expected**

$\hat y_{148} = b_0 +b_1x_{148}$

$\hat y_{148} = -10.1 + 0.464\times300$

$\hat y_{148}$ = `r -10.1 + 0.464*300`


]

.pull-left[

**Observed**

$y_{148} =$ 160

]

---

## Residual for `i = 148`

.pull-left[

```{r echo = FALSE, fig.align='center', message=FALSE, warning = FALSE, fig.height = 4}
babies %>% 
  ggplot() +
  aes(x = gestation, y = bwt) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = baby_148, color = "red") +
  geom_segment(x = 300, xend = 300, y = 128.94, yend = 160, color = "red")
```



]

.pull-right[

$y_{148} = 160$


<hr>

$\hat y_{148}$ = `r -10.1 + 0.464*300`


<hr>

$e_{148} = y_{148} - \hat y_{148}$

$e_{148} =$ `r 160 -(-10.1 + 0.464*300)`


]

---

## Least Squares Regression 

The goal is to minimize 

$$e_1^2 + e_2^2 + ... + e_n^2$$


