---
title: "Multiple Linear Regression"
author: "Created by Dr. Mine Dogucu and Presented by Dr. Jessica Jaynes"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    lib_dir: libs
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"


---

class: title-slide

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
]

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(gridExtra)
library(openintro)
library(broom)
options(scipen = 999)
```

---


class: middle

<div align = "center">

| y | Response    | Birth weight | Numeric |
|---|-------------|-----------------|---------|
| x | Explanatory | Smoke           | Categorical |


---

## Notation

$y_i = \beta_0 +\beta_1x_i + \epsilon_i$

$\beta_0$ is y-intercept  
$\beta_1$ is slope  
$\epsilon_i$ is error/residual  
$i = 1, 2, ...n$ identifier for each point


---

```{r}
model_s <- lm(bwt ~ smoke, data = babies)
tidy(model_s)
```



--

$\hat {y}_i = b_0 + b_1 x_i$

$\hat {\text{bwt}_i} = b_0 + b_1 \text{ smoke}_i$

$\hat {\text{bwt}_i} = 123 + (-8.94\text{ smoke}_i)$

---

class: middle

.pull-left[

#### Expected bwt for a baby with a non-smoker mother

$\hat {\text{bwt}_i} = 123 + (-8.94\text{ smoke}_i)$

$\hat {\text{bwt}_i} = 123 + (-8.94\times 0)$

$\hat {\text{bwt}_i} = 123$

$E[bwt_i | smoke_i = 0] = b_0$


]

--

.pull-right[
#### Expected bwt for a baby with a smoker mother

$\hat {\text{bwt}_i} = 123 + (-8.94\text{ smoke}_i)$

$\hat {\text{bwt}_i} = 123 + (-8.94\times 1)$

$\hat {\text{bwt}_i} = 114.06$

$E[bwt_i | smoke_i = 1] = b_0 + b_1$

]

---

```{r}
confint(model_s)
```

Note that the confidence interval for the "slope" does not contain 0 and all the values in the interval are negative. 

---

```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(openintro)
library(broom)
theme_set(theme_gray(base_size = 18))
```

#### Data `babies` in `openintro` package

```{r echo = FALSE}
glimpse(babies)
```

---


class: middle

<div align = "center">

| y | Response    | Birth weight | Numeric |
|---|-------------|-----------------|---------|
| $x_1$ | Explanatory | Gestation           | Numeric |
| $x_2$ | Explanatory | Smoke           | Categorical |

---

## Notation

$y_i = \beta_0 +\beta_1x_{1i}  + \beta_2x_{2i} + \epsilon_i$

$\beta_0$ is intercept  
$\beta_1$ is the slope for gestation   
$\beta_2$ is the slope for smoke 
$\epsilon_i$ is error/residual  
$i = 1, 2, ...n$ identifier for each point


---

```{r}
model_gs <- lm(bwt ~ gestation + smoke, data = babies)
tidy(model_gs)
```

--

Expected birth weight for a baby who had 280 days of gestation with a smoker mother

$\hat {\text{bwt}_i} = b_0 + b_1 \text{ gestation}_i + b_2 \text{ smoke}_i$

$\hat {\text{bwt}_i} = -0.932 + (0.443 \times 280) + (-8.09 \times 1)$

---

```{r}
babies %>% 
  modelr::add_predictions(model_gs) %>% 
  select(bwt, gestation, smoke, pred)
```

---

```{r}
babies %>% 
  modelr::add_predictions(model_gs) %>% 
  modelr::add_residuals(model_gs) %>% 
  select(bwt, gestation, smoke, pred, resid)
```
---

---
# Collinearity between explanatory variables
- Two predictor variables are said to be collinear when they are  correlated, and this collinearity complicates model estimation.
Remember: Predictors are also called explanatory or independent variables. Ideally,  they would be independent of each other.

- We don’t like adding predictors that are associated with each  other to the model, because often times the addition of such  variable brings nothing to the table. Instead, we prefer the  simplest best model, i.e. parsimonious model.

- While it’s impossible to avoid collinearity from arising in observational data, experiments are usually designed to prevent correlation among predictors.

---
Great example of simple linear regression in action!
https://www.tandfonline.com/doi/full/10.1080/26939169.2021.1946450



