---
title: "Introduction"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    lib_dir: libs
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"


---

class: title-slide

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
]
---
class: middle center

.font150[Exploring Data]


---

class: middle
```{r message = FALSE}
library(tidyverse)
library(openintro)

```

```{r echo = FALSE}
fastfood <- fastfood %>% 
  mutate(restaurant = as.factor(restaurant))
```



```{r}
glimpse(fastfood)
```



---
class: middle

## Review

**mean** : average of a list of data values 
$$\bar x = \frac{\Sigma_{i = 1}^{n} x_i}{n}$$

**median**: the middle value when the data are ordered

---

**sample variance**: average square distance from the mean. It measures variability of the data.
$$s^2 = \frac{\Sigma_{i = 1}^{n} (x_i - \bar x )^2}{n-1}$$


**sample standard deviation**: average distance of a data value from the mean. Square root of the variance. 

$$s = \sqrt{s^{2}} = \sqrt{\frac{\Sigma_{i = 1}^{n} (x_i - \bar x )^2}{n-1}}$$


---
## Mean
**mean** : average of a list of data values 
$$\bar x = \frac{\Sigma_{i = 1}^{n} x_i}{n}$$
--
.pull-left[
```{r}
summarize(fastfood, 
          mean(calories))
```
]

--

.pull-right[

```{r}
mean(fastfood$calories)
```

]


---
## Median
**median**: the middle value when the data are ordered

--
.pull-left[
```{r}
summarize(fastfood, 
          median(calories))
```
]

--

.pull-right[

```{r}
median(fastfood$calories)
```

]



---
## Variance
**sample variance**: average square distance from the mean. It measures variability of the data.
$$s^2 = \frac{\Sigma_{i = 1}^{n} (x_i - \bar x )^2}{n-1}$$
--

.pull-left[
```{r}
summarize(fastfood, 
          var(calories))
```
]

--

.pull-right[

```{r}
var(fastfood$calories)
```

]


---
## Standard Deviation
**sample standard deviation**: average distance of a data value from the mean. Square root of the variance. 
$$s = \sqrt{s^{2}} = \sqrt{\frac{\Sigma_{i = 1}^{n} (x_i - \bar x )^2}{n-1}}$$
--

.pull-left[
```{r}
summarize(fastfood, 
          sd(calories))
```
]

--

.pull-right[

```{r}
sd(fastfood$calories)
```

]


---
## Minimum

.pull-left[
```{r}
summarize(fastfood, 
          min(calories))
```
]

--

.pull-right[

```{r}
min(fastfood$calories)
```

]



---
## Maximum

.pull-left[
```{r}
summarize(fastfood, 
          max(calories))
```
]

--

.pull-right[

```{r}
max(fastfood$calories)
```

]



---
We can use multiple functions inside `summarize`. 

```{r}
summarize(fastfood, 
          mean(calories), 
          median(calories),
          var(calories),
          sd(calories),
          min(calories),
          max(calories))
```

---

```{r}
fastfood %>% 
  group_by(restaurant) %>% 
  summarize(med_calories = median(calories),
            mean_calories = mean(calories))
```


