---
title: "Changing Variables"
author: "Dr. Mine Dogucu"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    lib_dir: libs
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"


---

class: title-slide

<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`
]

---

class: center middle

.font150[Review]

---

class: middle

How many panes have we seen in RStudio and what is the purpose of each pane?

---

class: middle

Which of the following files is a markdown file?

a. `example.R`  
b. `example.md`  
c. `example.Rmd`  

---


class: middle

Which of the following is a valid order of actions when starting a project using git and GitHub?

a. clone, commit, push, 

b. push, commit, clone

c. commit, clone, push

d. clone, push, commit

---

class: middle

Which R functions have we learned together?

---

class: middle

What is the formula for variance?

---

class: middle

You are given a data frame called `registrar`. There are two variables you are interested in `class_year` which represents whether someone is a first year, sophomore, junior, or senior and `gpa` which represents GPA. 

How would you find the average GPA for each class rank?

---


class: center middle

.font150[Changing Variables]


---


class: middle

```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(janitor)
options(scipen = 999)
```



```{r echo = FALSE, message = FALSE}
lapd <- read_csv(here::here("slides/data/Police_Payroll.csv")) %>% 
  select(2, 7, 17)
```


```{r}
glimpse(lapd)
```

---

class: middle

`clean_names()` makes variable names in tidy style.

```{r}

lapd <- clean_names(lapd)
glimpse(lapd)
```


---

class: middle

**Goal**: 

Create a new variable called `base_pay_k` that represents `base_pay` in thousand dollars.

---

class: middle

```{r}
lapd %>% 
  mutate(base_pay_k = base_pay/1000)
```

---

class: middle

```{r}
glimpse(lapd)
```

**Goal**: 

Create a new variable called `base_pay_level` which has `Less Than 0`, `No Income`, `Less than Median and Greater than 0` and `Greater than Median`. We will consider $62474 as the median (from previous lecture).

---


```{r}
lapd %>% 
  mutate(base_pay_level = case_when(
    base_pay < 0 ~ "Less than 0", 
    base_pay == 0 ~ "No Income",
    base_pay < 62474 & base_pay > 0 ~ "Less than Median, Greater than 0",
    base_pay > 62474 ~ "Greater than Median")) 
```


---

class: middle

## (Some) Variable Types in R

`character`: takes string values (e.g. a person's name, address)  
`integer`: integer (single precision)  
`double`: floating decimal (double precision)  
`numeric`: integer or double  
`factor`: categorical variables with different levels  
`logical`: TRUE (1), FALSE (0)  

---

class: middle

```{r}
glimpse(lapd)
```

**Goal**: 

Change `year` and `employment_type` to appropriate variable types. 


---

class: middle

```{r}
lapd %>% 
  mutate(employment_type = as.factor(employment_type),
         year = as.integer(year)) 
```

---

class: middle

`as.factor()` - makes a vector factor  
`as.numeric()` - makes a vector numeric  
`as.integer()` - makes a vector integer  
`as.double()` - makes a vector double  
`as.character()` - makes a vector character  

---


In your lecture notes, you can do all the changes in this lecture in one long set of piped code. That's the beauty of piping!

```{r eval = FALSE}
lapd <- 
  lapd %>% 
  clean_names() %>% 
    mutate(employment_type = as.factor(employment_type),
           year = as.integer(year),
           base_pay_level = case_when(
             base_pay < 0 ~ "Less than 0", 
             base_pay == 0 ~ "No Income",
             base_pay < 62474 & base_pay > 0 ~ "Less than Median, Greater than 0",
             base_pay > 62474 ~ "Greater than Median")) 
```

---

class: middle

## Word of caution

The functions `clean_names()`, and `mutate()` all take a data frame as the first argument. Even though we do not see it, the data frame is piped through from the previous step of code at each step. 
When we use these functions without the `%>%` we have to include the data frame explicitly.


.pull-left[
Data frame is used as the first argument
```{r}
clean_names(lapd)
```
]

.pull-right[
Data frame is piped

```{r}
lapd %>% 
  clean_names()
```
]